Estimating survival in left filtered data

### 2021-12-14 paper.tex
Revised draft

The United Auto Workers-General Motors (UAW-GM) Cohort Study is a
longitudinal occupational cohort study established in the early 1980s to
study the health effects of metalworking fluids (Eisen et al. 1992,
2001). Metalworking fluids (MWF) are complex mixtures of fluids used in
industrial metalworking operations to lubricate and cool machinery and
parts. The three major classes of MWF are straight, soluble, and
synthetic metalworking fluids (Byers 2006). Possible routes of human
exposure include absorption through skin, inhalation or aerosols, and
ingestion of droplets.

A central concern in the analysis of occupational cohorts is the
potential for the healthy worker survivor effect (HWSE), the phenomenon
by which healthy individuals remain at work, while less healthy
individuals leave work -- possibly in response to exposure-related
health decline. In the presence of the HWSE, those with the highest
cumulative occupational exposures are also those who are less at risk of
disease. Thus, standard measures of association would show an inverse
relationship between occupational exposure and poor health outcomes
(Arrighi and Hertz-Picciotto 1994). The HWSE is an example of
time-varying confounding affected by past exposure. Previous studies
have attempted to assess the presence of the HWSE in observed data by
assessing so-called path-specific associations using Cox proportional
hazards modeling (Naimi et al. 2013; Garcia et al. 2017). However, these
measures of associations were themselves subject to the confounding
structures they sought to characterize.

If conditional sequential ignorability of exposure and censoring status
at each point in follow-up and positivity can be attained given
covariate history, then causal methods can be applied to account for the
HWSE. Past studies have applied causal methods capable of accounting for
time-varying confounding affected by past exposure to the study of MWF
exposures and cancer mortality outcomes in the UAW-GM Cohort Study
(Garcia et al. 2018; Izano et al. 2019), but the study of cancer
incidence outcomes is further problematized by the incomplete
observation of cancer incidence outcomes at every point of follow-up
over the study period. Nonetheless, we wish to make inferences about the
carcinogenicity of MWF exposure over an individual's lifetime starting
upon entry into the workforce. The UAW-GM Cohort included those hired
roughly between 1938 and 1985. However, cancer incidence reporting at
the Michigan Cancer Registry did not begin until 1985. Hence, our
observed cohort data is subject to \emph{left filtering}: cancer
incidence is the outcome of interest, but before 1985, both cancer
incidence status and time of cancer incidence are unknown. Observation
of the complete cancer incidence outcome vector over the study period
was conditional on an individual surviving to 1985 cancer-free.

In the presence of the HWSE, left filtering implies outcome
misclassification that is informative of true cancer status. As part of
her dissertation research, Izano (2017) conducted a quantitative bias
analysis for the estimation of survival curves using left filtered data
in the presence of HWSE. She simulated data compatible with the HWSE and
estimated cancer-free survival curves using an inverse probability of
treatment and censoring weighted Kaplan-Meier (WKM) estimator and the
WKM with an Aalen filter for left-filtering (AWKM) (Andersen et al.
1993; Xie and Liu 2005). Data were simulated under five different
scenarios. The survival curves under the following interventions were
computed or estimated in 250 replicates: (1) always exposed at work with
no censoring due to death and (2) never exposed at work with no
censoring due to death. For each intervention, three survival curves
were produced: (1) the true survival curve, (2) the WKM survival curve,
and (3) AWKM survival curve. Estimator bias was evaluated by comparing
the average of 250 estimates of to the average difference in cancer-free
survival under the two rules to the true average survival difference
when exposure and censoring were controlled deterministically.

The present project replicates the simulation and bias analyses
presented in Chapter 3 of Izano (2017), embeds the problem in the
non-parametric structural causal approach of Pearl (1995), and applies
the WKM and AWKM estimators to the estimation of digestive-system
cancer-free survival under exposure/censoring rules in the UAW-GM
Cohort. The \texttt{stremr} package was used to estimate the WKM and
AWKM survival curves in both the simulation and the applied analyses
(Sofrygin, van der Laan, and Neugebauer 2021).

@@

The UAW-GM Cohort data included person-year level exposure, outcome, and
covariate data starting at hire. To emulate the shape of the data for
this longitudinal cohort, we considered 20 years of data over time
indexed by years since hire. Notation representing the variables of
interest are presented in Table \ref{tab:variables}. The causal model
represents hypothetical relationships between variables over time
compatible with the theory underlying the HWSE in longitudinal
occupational cohort studies. At each time point, the effect of
cumulative exposure \(\bar A(t)\) on cancer incidence \(Y^(t)\) is
confounded by a path through employment status \(N(t)\) and underlying
health \(H(t)\) as well as the path through past exposure
\(\bar A(t - 1)\) and vital status \(D(t)\). These paths follow
straightforward logic: occupational exposure depends upon employment
status and past exposure; mortality status is affected by past exposure
and cancer history. Confounding by baseline covariates \(W\) is assumed
throughout.

@@

Assume we have \(n = 50\,000\) iid units in \(X\) with
\[X_i(t) = \left(R_i = 0, W_i, S_i, \bar H_i(t), \bar N_i(t), \bar A_i(t), \bar a_i(t), \bar Y^*_i(t) = \bar Y_i(t) \right).\]
The bar notation to indicates variable history where
\(\bar X_i(t) = \left(X_i(k)\right)_{k = 1}^t\). In the case of
exposure, \(\bar A_i(t)\) is the indicator of whether unit \(i\) was
ever exposed through time \(t\). Note that true cancer status \(Y^*(t)\)
not observed until \(t \ge R\), after the start of the registry. Call
\(X\) the full data, where we have \(R = 0\) for all. In the observed
data \(X^\text{obs}\), we cannot assume \(R = 0\) for all;
susceptibility \(S\) and underlying health status \(H\) are not known:
\[X^\text{obs}_i(t) = \left(R_i, W_i, \bar N_i(t), \bar A_i(t), \bar a_i(t), \bar Y_i(t) \right).\]
Under the causal model, we assume the following non-parametric
structural equations: \[\begin{aligned}
R     & = f_R \left( U_R \right) \\
W     & = f_W \left( U_W \right) \\
S     & = f_S \left( U_S \right) \\
H(t)  & = f_{H(t)} \left( H(t - 1), U_{H(t)} \right) \\
N(t)  & = f_{N(t)} \left( W, N(t - 1), H(t), A(t - 1), U_{N(t)} \right) \\
A(t)  & = f_{A(t)} \left( W, \bar A(t - 1), N(t), U_{A(t)} \right) \\
D(t)  & = f_{D(t)} \left( W, \bar A(t - 1), D(t - 1), Y^*(t - 1), U_{D(t)} \right) \\
Y^*(t)& = f_{Y^*(t)} \left(W, S, H(t), \bar A (t), D(t), Y^*(t - 1), U_{Y^*(t)} \right) \\
Y(t) & = Y^*(t) \times \Ind{Y^*(\lfloor R \rfloor) = 0} \times \Ind{D(t) = 0} \ .
\end{aligned}\] The exogenous variables (errors)
\(U = \left( U_R, U_W, U_S, U_{H(t)}, U_{N(t)}, U_{A(t)}, U_{D(t)}, U_{Y^*(t)} \right)_{t = 1}^{T}\)
are mutually independent. Exposure status is a time-varying indicator;
let exposure history \(\bar A(t)\) be summarized as being ever-exposed
or not with \(\bar A(t) = \Ind{\sum_{k = 1}^t\Ind{A(k) = 1} > 0}\). The
outcome of interest is a survival outcome, so
\(Y^*(t - 1) = 1 \Rightarrow Y^*(t) = 1\). The observed outcome \(Y(t)\)
at time \(t\) is a function of true outcome status and the time points
delimiting left and right censoring. An abbreviated directed acyclic
graph (DAG) representing the causal relationships encoded in the
equations above is presented in Figure \ref{fig:dag}.

\begin{figure}
\caption{Directed acyclic graph representing the causal relationships encoded in the non-parametric structural equation model at time $t$.}
\label{fig:dag}
\begin{center}
\begin{tikzpicture}[>= stealth, auto, node distance = 2.25cm, semithick, fill=white, inner sep=0pt]
\tikzstyle{every state}=[shape = circle, align = center, draw = none]
\node[state] (H0) {$H(t - 1)$};
\node[state] (N0) [below right of=H0] {$N(t - 1)$};
\node[state] (A0) [below right of=N0] {$\bar A(t - 1)$};
\node[state] (D0) [below right of=A0] {$D(t - 1)$};
\node[state] (Y0) [below right of=D0] {$Y^*(t - 1)$};
\node[state] (W) [above right of=H0] {$W$};
\node[state] (H) [right of=H0, node distance = 4cm] {$H(t)$};
\node[state] (N) [below right of=H, node distance = 4cm] {$N(t)$};
\node[state] (A) [below right of=N, node distance = 4cm] {$\bar A(t)$};
\node[state] (D) [above right of=N, node distance = 3cm] {$D(t)$};
\node[state] (Y) [above right of=A, node distance = 3cm] {$Y^*(t)$};
\node[state] (S) [above right of=D, node distance = 3cm] {$S$};
%
\path[->] (H0) edge (N0);
\path[->] (N0) edge (A0);
\path[->] (A0) edge [bend right=20] (Y0);
\path[->] (H0) edge (H);
\path[->] (W) edge [bend right=10] (N);
\path[->] (N0) edge (N);
\path[->] (H) edge (N);
\path[->] (A0) edge (N);
\path[->] (W) edge [bend right=15] (A);
\path[->] (A0) edge (A);
\path[->] (N) edge (A);
\path[->] (W) edge (D);
\path[->] (A0) edge [bend left=15] (D);
\path[->] (D0) edge [bend right=15] (D);
\path[->] (Y0) edge [bend right=10] (D);
\path[->] (W) edge [bend left=5] (Y);
\path[->] (S) edge [bend left=30] (Y);
\path[->] (S) edge [bend left=15] (Y0);
\path[->] (H) edge (Y);
\path[->] (A) edge (Y);
\path[->] (D) edge (Y);
\path[->] (Y0) edge (Y);
\end{tikzpicture}
\end{center}
\end{figure}

Figure \ref{fig:dag} clarifies our conceptualization of the HWSE. At
each time point \(t\), the cumulative effect of exposure \(\bar A(t)\)
on cancer incidence \(Y^*(t)\) is confounded by employment status
\(N(t)\) through the backdoor path
\(\bar A(t) \leftarrow N(t) \leftarrow H(t) \rightarrow Y^*(t)\). In the
absence of observed data on health status \(H(t)\), an analyst may be
tempted to condition on employment status \(N(t)\), but doing so would
introduce collider bias while blocking the causal path between past
exposure \(\bar A(t - 1)\) and cancer \(Y^*(t)\). Furthermore, an
analysis starting at an arbitrary time point after the time origin
(hire) would be tantamount to conditioning on those still alive at that
time, which would also result in both collider bias and the conditioning
on nodes on the causal path between the exposure and outcome of
interest.

@@

Five sets of data were generated using these equations to represent five
scenarios. Scenario 1 represents the base case where 10\% of workers are
susceptible to exposure-related effects, the odds ratio of mortality
each additional year following cancer diagnosis is about 1.6, and there
is moderate time-varying confounding by health status. In scenario 2, we
have greater cancer-related mortality by increasing
\(\beta_{\bar Y}^D\). In scenario 3, we increase \(p_S\), the proportion
of the study population susceptible to the carcinogenic effects of MWF
exposure. In scenario 4, we consider greater time-varying confounding by
health status by increasing \(\beta_H^N\) and \(\beta_H^Y\). In the last
scenario, we have greater background cancer incidence by increasing
\(\beta_0^Y\). The sets of parameters used in the five scenarios are
presented in Table \ref{tab:params}.

@@

The substantive question of interest was the causal effect of
occupational exposure to MWF on cancer incidence risk. Since
occupational MWF exposure occurs only when individuals are at work, we
defined dynamic exposure regimes that depend on employment status. Under
rule \(a_0\), set \(D(t) = 0\), and set \(A(t) = 0\) while \(N(t) = 1\).
Under rule \(a_1\), set \(D(t) = 0\), and set \(A(t) = 1\) while
\(N(t) = 1\). Under both rules, we prevented censoring by death as if it
were intervenable. The causal effect was defined by contrasting the
survival function \(S_{a_1}(t) = 1 - \E{Y_{a_1}(t)}\) under rule \(a_1\)
to \(S_{a_0}(t) = 1 - \E{Y_{a_0}(t)}\) that under rule \(a_0\). Note
that this causal estimand was defined over \emph{a priori
counterfactuals} not observable in the real world (Frangakis and Rubin
2002). This approach is standard in epidemiologic studies, however.

The survival function expresses the probability that a person following
rule \(a\) is cancer-free at the end of time point \(t\). The expected
time until cancer under rule \(a\) is
\(\mu_{a} = \sum_0^K S_{a}(t) \, dt\). The parameter used in the
estimation of bias was the summary measure
\(\psi = \mu_{a_1} - \mu_{a_0}\), the difference in expected time until
event under two different interventions over 20 years of follow-up under
five different data generating scenarios. Bias was evaluated by
comparing estimates of \(\psi\) to its true value in 250 simulations per
scenario (the original analysis performed 500). The true value was
calculated by simulating the full data for five hundred thousand
individuals (the original analysis used one million) with rules \(a_0\)
and \(a_1\) applied deterministically. Estimates of \(\psi\) were
obtained by first estimating the survival curves \(S_a(t)\) using two
estimators: the inverse probability weighted Kaplan-Meier estimator
(WKM) and the Aalen-filtered WKM (AWKM). These survival estimators are
detailed in the following section.

@@

To estimate survival, we applied extensions of the widely-known
Kaplan-Meier (KM) estimator for survival (Kaplan and Meier 1958). First,
we review the estimator of Xie and Liu (2005), an extension of the KM
estimator where units are weighted by the inverse probability of
treatment. The standard KM estimator requires counting up the number of
cases \(c^0_{a}(t)\) that occurred in interval \((t - 1, t]\) and the
number of units at risk \(R^0_{a}(t)\) in that interval at all event
times \(t\). Assuming cancer status was assessed at the end of regular
intervals \(t = 1, \ldots, K\), we have: \[\begin{aligned}
c^0_{a} (t)
& = \sum_i^n \Ind{Y_i(t) = 1} \times \Ind{Y_i(t - 1) = 0} \times \Ind{\bar A_i (t) = \bar a(t)} \\
R^0_{a} (t)
& = \sum_i^n \Ind{Y_i(t - 1) = 0} \times \Ind{\bar A_i (t) = \bar a(t)}.
\end{aligned}\] The standard survival estimator is \[\hat S^0_{a}(t) =
\begin{cases}
1 & \text{if } t < t_1 \\
\prod_{j \le t} \left(1 - \frac{c^0_{a}(j)}{R^0_{a}(j)}\right) & \text{if } t \ge t_1 \\
\end{cases}\] where \(t_1\) is the first event time.

In observational studies, survival contrasts estimated using the
standard KM estimator are biased for the true causal survival contrast.
However, if conditional ignorability and positivity are attained, the
inverse probability weighted KM (WKM) estimator of Xie and Liu (2005)
yields unbiased estimates of the true causal survival curve. The WKM
estimator augments the standard KM estimator by weighting units at time
\(t\) by \(w_{i, a}(t)\) the inverse probability of treatment:
\[\begin{aligned}
c^w_{a} (t)
& = \sum_i^n w_{i, a}(t) \times \Ind{Y_i(t) = 1} \times \Ind{Y_i(t - 1) = 0} \times \Ind{\bar A_i (t) = \bar a(t)}\\
R^w_{a} (t)
& = \sum_i^n w_{i, a}(t) \times \Ind{Y_i(t - 1) = 0} \times \Ind{\bar A_i (t) = \bar a(t)}
\end{aligned}\] The WKM survival estimator for rule \(a\) is
\[\hat S^w_{a}(t) =
\begin{cases}
1 & \text{if } t < t_1 \\
\prod_{j \le t} \left(1 - \frac{c^w_{a}(j)}{R^w_{a}(j)}\right) & \text{if } t \ge t_1 \\
\end{cases}\] where \(t_1\) is the first event time.

Finally, to account for (uninformative) left filtering, we applied the
Aalen filter, which considers only the units at time \(t\) for which the
outcome is observed: \[\begin{aligned}
c_{a} (t)
& = \sum_i^n w_{i, a}(t) \times \Ind{Y_i(t) = 1} \times \Ind{Y_i(t - 1) = 0} \times \Ind{\bar A_i (t) = \bar a(t)} \times \Ind{t \ge R_i} \\
R_{a} (t)
& = \sum_i^n w_{i, a}(t) \times \Ind{Y_i(t - 1) = 0} \times \Ind{\bar A_i (t) = \bar a(t)} \times \Ind{t \ge R_i}
\end{aligned}\] The Aalen-filtered WKM (AWKM) estimator for rule \(a\)
is \[\hat S_{a}(t) =
\begin{cases}
1 & \text{if } t < t_1 \\
\prod_{j \le t} \left(1 - \frac{c_{a}(j)}{R_{a}(j)}\right) & \text{if } t \ge t_1 \\
\end{cases}\] where \(t_1\) is the first event time.

In the full data, the WKM and AWKM estimators are equivalent, and
identification is achieved under overlap (positivity) and sequential
ignorability (randomization) assumptions: \[\begin{aligned}
Y^*_{a, \bar d = 0} (t')   & \indep A(t) \mid W,\ \bar A(t - 1) = \bar a(t - 1),\ D(t - 1) = 0,\ N(t) = 1\\
Y^*_{a, \bar d = 0} (t') & \indep D(t) \mid W,\ D(t - 1) = 0,\ Y^*(t - 1) = 0,\ \bar A(t - 1) = \bar a(t - 1)
\end{aligned}\] for all times \(t'\ge t\), and
\[0 < \Prob{ A(t) = 1 \mid W,\ \bar A(t - 1) = \bar a(t - 1),\ D(t - 1) = 0,\ N(t) = 1} < 1\]
\[0 < \Prob{D(t) = 0 \mid W,\ D(t - 1) = 0,\ Y^*(t - 1) = 0,\ \bar A(t - 1) = \bar a(t - 1)} < 1.\]

>>>>

The United Auto Workers-General Motors (UAW-GM) Cohort Study is a
longitudinal occupational cohort study established in the early 1980s to
study the health effects of metalworking fluids (Eisen et al. 1992,
2001). Metalworking fluids (MWF) are complex mixtures of fluids used in
industrial metalworking operations to lubricate and cool machinery and
parts. The three major classes of MWF are straight, soluble, and
synthetic metalworking fluids (Byers 2006). Possible routes of human
exposure include absorption through skin, inhalation of aerosols, and
ingestion of droplets.

A central concern in the analysis of longitudinal occupational cohort
data is the potential for bias due to the healthy worker survivor effect
(HWSE), the phenomenon by which healthy individuals remain at work,
while less healthy individuals leave work -- possibly in response to
exposure-related health decline. In the presence of the HWSE, those with
the highest cumulative occupational exposures are also those who are
less at risk of disease (more likely to have potential outcome no
disease under both exposure and no exposure scenarios). Thus, standard
measures of association would show an inverse relationship between
occupational exposure and poor health outcomes (Arrighi and
Hertz-Picciotto 1994). The HWSE is an example of time-varying
confounding affected by past exposure. Previous studies have attempted
to assess the presence of the HWSE in observed data by assessing
so-called path-specific associations using Cox proportional hazards
modeling (Naimi et al. 2013; Garcia et al. 2017). However, in those
studies, the estimates were themselves subject to the confounding
structures they sought to characterize.

If conditional sequential ignorability of exposure and censoring status
at each point in follow-up and positivity (overlap) can be attained
given covariate history, then longitudinal causal methods can be applied
to account for the HWSE. Past studies have applied causal methods
capable of accounting for time-varying confounding affected by past
exposure to the study of MWF exposures and cancer mortality outcomes in
the UAW-GM Cohort Study (Garcia et al. 2018; Izano et al. 2019).
However, the study of cancer incidence outcomes is further problematized
by the incomplete observation of cancer incidence outcomes at every
point of follow-up over the study period. Nonetheless, we wish to make
inferences about the carcinogenicity of MWF exposure over an
individual's lifetime starting upon entry into the workforce. The UAW-GM
Cohort included those hired roughly between 1938 and 1985. However,
cancer incidence reporting did not begin until 1973. Hence, our observed
cohort data exhibits \emph{left filtering}: cancer incidence is the
outcome of interest, but before 1985, both cancer incidence status and
time of cancer incidence are unknown. Observation of the complete cancer
incidence outcome vector over the study period was conditional on an
individual surviving to 1985 cancer-free.

In the presence of the HWSE, left filtering implies outcome
misclassification that is informative of true cancer status. As part of
her dissertation research, Izano (2017) conducted a quantitative bias
analysis for the estimation of survival curves using left filtered data
in the presence of HWSE. She simulated data compatible with the HWSE and
estimated cancer-free survival curves using an inverse probability of
treatment and censoring weighted Kaplan-Meier (WKM) estimator and the
WKM with an Aalen filter for left-filtering (AWKM) (Andersen et al.
1993; Xie and Liu 2005). Data were simulated under five different
scenarios. The survival curves under the following interventions were
computed or estimated in 250 replicates: (1) always exposed at work with
no censoring due to death and (2) never exposed at work with no
censoring due to death. For each intervention, three survival curves
were produced: (1) the true survival curve, (2) the WKM survival curve,
and (3) AWKM survival curve. Estimator bias was evaluated by comparing
the average of 250 estimates of to the average difference in cancer-free
survival under the two rules to the true average difference in survival
when exposure and censoring were controlled deterministically.

The present project replicates the simulation and bias analyses
presented in Chapter 3 of Izano (2017), embeds the problem in the
non-parametric structural causal approach of Pearl (1995), and applies
the WKM and AWKM estimators to the estimation of digestive-system
cancer-free survival under exposure and no censoring rules in the UAW-GM
Cohort. The \texttt{stremr} package was used to estimate the WKM and
AWKM survival curves in both the simulation and the applied analyses
(Sofrygin, van der Laan, and Neugebauer 2021).

@@

The UAW-GM Cohort data included person-year level exposure, outcome, and
covariate data starting at hire. To emulate the shape of the data for
this longitudinal cohort, we considered 20 years of data over time
indexed by years since hire. Notation representing the variables of
interest are presented in Table \ref{tab:variables}. The causal model
represents hypothetical relationships between variables over time
compatible with the theory underlying the HWSE in longitudinal
occupational cohort studies. At each time point, the effect of
cumulative exposure \(\bar A(t)\) on cancer incidence \(Y^*(t)\) is
confounded by the path through employment status \(N(t)\) and underlying
health \(H(t)\) as well as the path through past exposure
\(\bar A(t - 1)\) and vital status \(D(t)\). These paths follow
straightforward logic: occupational exposure depends upon employment
status and past exposure; mortality status is affected by past exposure
and cancer history. Confounding by baseline covariates \(W\) is assumed
throughout.

@@

Assume we have \(n = 50\,000\) iid units in \(X\) with
\[X_i(t) = \left(R_i = 0, W_i, S_i, \bar H_i(t), \bar N_i(t), \bar A_i(t), \bar a_i(t), \bar Y^*_i(t) = \bar Y_i(t) \right).\]
In general, we use bar notation to indicate variable history as follows
\(\bar X_i(t) = \left(X_i(k)\right)_{k = 1}^t\). Note that true cancer
status \(Y^*(t)\) is not observed until \(t \ge R\), after the start of
the registry. Call \(X\) the full data, where we have \(R = 0\) for all.
In the observed data \(X^\text{obs}\), we cannot assume \(R = 0\) for
all. Additionally, susceptibility \(S\) and underlying health status
\(H\) are not known:
\[X^\text{obs}_i(t) = \left(R_i, W_i, \bar N_i(t), \bar A_i(t), \bar a_i(t), \bar Y_i(t) \right).\]
Under the causal model, we assume the following non-parametric
structural equations: \[\begin{aligned}
R     & = f_R \left( U_R \right) \\
W     & = f_W \left( U_W \right) \\
S     & = f_S \left( U_S \right) \\
H(t)  & = f_{H(t)} \left( H(t - 1), U_{H(t)} \right) \\
N(t)  & = f_{N(t)} \left( W, N(t - 1), H(t), A(t - 1), U_{N(t)} \right) \\
A(t)  & = f_{A(t)} \left( W, \bar A(t - 1), N(t), U_{A(t)} \right) \\
D(t)  & = f_{D(t)} \left( W, \bar A(t - 1), D(t - 1), Y^*(t - 1), U_{D(t)} \right) \\
Y^*(t)& = f_{Y^*(t)} \left(W, S, H(t), \bar A (t), D(t), Y^*(t - 1), U_{Y^*(t)} \right) \\
Y(t) & = Y^*(t) \times \Ind{Y^*(\lfloor R \rfloor) = 0} \times \Ind{D(t) = 0} \ .
\end{aligned}\] The exogenous variables (errors)
\(U = \left( U_R, U_W, U_S, U_{H(t)}, U_{N(t)}, U_{A(t)}, U_{D(t)}, U_{Y^*(t)} \right)_{t = 1}^{T}\)
are mutually independent. Exposure status is a time-varying indicator,
and exposure history is summarized as
\(\bar A(t) = \Ind{\sum_{k = 1}^t\Ind{A(k) = 1} > 0}\). The outcome of
interest is a survival outcome, so
\(Y^*(t - 1) = 1 \Rightarrow Y^*(t) = 1\). The observed outcome \(Y(t)\)
at time \(t\) is a function of true outcome status, time of left
censoring, and time of right censoring. An abbreviated directed acyclic
graph (DAG) representing the causal relationships encoded in the
equations above is presented in Figure \ref{fig:dag}.

\begin{figure}
\caption{Directed acyclic graph representing the causal relationships encoded in the non-parametric structural equation model at time $t$.}
\label{fig:dag}
\begin{center}
\begin{tikzpicture}[>= stealth, auto, node distance = 2.25cm, semithick, fill=white, inner sep=0pt]
\tikzstyle{every state}=[shape = circle, align = center, draw = none]
\node[state] (H0) {$H(t - 1)$};
\node[state] (N0) [below right of=H0] {$N(t - 1)$};
\node[state] (A0) [below right of=N0] {$\bar A(t - 1)$};
\node[state] (D0) [below right of=A0] {$D(t - 1)$};
\node[state] (Y0) [below right of=D0] {$Y^*(t - 1)$};
\node[state] (W) [above right of=H0] {$W$};
\node[state] (H) [right of=H0, node distance = 4cm] {$H(t)$};
\node[state] (N) [below right of=H, node distance = 4cm] {$N(t)$};
\node[state] (A) [below right of=N, node distance = 4cm] {$\bar A(t)$};
\node[state] (D) [above right of=N, node distance = 3cm] {$D(t)$};
\node[state] (Y) [above right of=A, node distance = 3cm] {$Y^*(t)$};
\node[state] (S) [above right of=D, node distance = 3cm] {$S$};
%
\path[->] (H0) edge (N0);
\path[->] (N0) edge (A0);
\path[->] (A0) edge [bend right=30] (Y0);
\path[->] (H0) edge (H);
\path[->] (W) edge [bend right=10] (N);
\path[->] (N0) edge (N);
\path[->] (H) edge (N);
\path[->] (A0) edge (N);
\path[->] (W) edge [bend right=15] (A);
\path[->] (A0) edge (A);
\path[->] (N) edge (A);
\path[->] (W) edge (D);
\path[->] (A0) edge [bend left=15] (D);
\path[->] (D0) edge [bend right=15] (D);
\path[->] (Y0) edge [bend right=10] (D);
\path[->] (W) edge [bend left=5] (Y);
\path[->] (S) edge [bend left=30] (Y);
\path[->] (S) edge [bend left=15] (Y0);
\path[->] (H) edge (Y);
\path[->] (A) edge (Y);
\path[->] (D) edge (Y);
\path[->] (Y0) edge (Y);
\end{tikzpicture}
\end{center}
\end{figure}

Figure \ref{fig:dag} clarifies our conceptualization of the HWSE. At
each time point \(t\), the cumulative effect of exposure \(\bar A(t)\)
on cancer incidence \(Y^*(t)\) is confounded by employment status
\(N(t)\) through the backdoor path
\(\bar A(t) \leftarrow N(t) \leftarrow H(t) \rightarrow Y^*(t)\). In the
absence of observed data on health status \(H(t)\), an analyst may be
tempted to conduct an analysis by simply ``blocking'' or conditioning on
employment status \(N(t)\), but doing so would introduce collider bias
while blocking the causal path between past exposure \(\bar A(t - 1)\)
and cancer \(Y^*(t)\). Furthermore, an analysis starting at an arbitrary
time point after the time origin (hire) would be tantamount to
conditioning on those still alive at that time, which would result in
both collider bias and the conditioning on nodes on the causal path
between the exposure and the outcome of interest.

@@

Five sets of data were generated using these equations, but with
different parameters, to represent five scenarios. Scenario 1 represents
the base case where 10\% of workers are susceptible to exposure-related
effects, the odds ratio of mortality each additional year following
cancer diagnosis is about 1.6, and there is moderate time-varying
confounding by health status. In scenario 2, we have greater
cancer-related mortality by increasing \(\beta_{\bar Y}^D\). In scenario
3, we increase \(p_S\), the proportion of the study population
susceptible to the carcinogenic effects of MWF exposure. In scenario 4,
we consider greater time-varying confounding by health status by
increasing \(\beta_H^N\) and \(\beta_H^Y\). In the last scenario, we
have greater background cancer incidence by increasing \(\beta_0^Y\).
The sets of parameters used in the five scenarios are presented in Table
\ref{tab:params}.

@@

The substantive question of interest was the causal effect of
occupational exposure to MWF on cancer incidence risk. Since
occupational MWF exposure occurs only when individuals are at work, we
defined dynamic exposure regimes that depend on employment status. Under
rule \(a_0\), set \(D(t) = 0\), and set \(A(t) = 0\) while \(N(t) = 1\).
Under rule \(a_1\), set \(D(t) = 0\), and set \(A(t) = 1\) while
\(N(t) = 1\). Under both rules, we prevented censoring by death as if it
were intervenable. The causal effect was defined by contrasting the
survival functions \(S_{a_1}(t) = 1 - \E{Y_{a_1}(t)}\) under rule
\(a_1\) to \(S_{a_0}(t) = 1 - \E{Y_{a_0}(t)}\) that under rule \(a_0\).
Note that this causal estimand was defined over \emph{a priori
counterfactuals} not observable in the real world (Frangakis and Rubin
2002). This approach is standard in epidemiologic studies.

The survival function expresses the probability that a person following
rule \(a\) is cancer-free at the end of time point \(t\). The expected
time until cancer under rule \(a\) is
\(\mu_{a} = \sum_0^K S_{a}(t) \, dt\). The parameter used in the
estimation of bias was the summary measure
\(\psi = \mu_{a_1} - \mu_{a_0}\), the difference in expected time until
event under two different interventions over 20 years of follow-up under
five different data generating scenarios. Bias was evaluated by
comparing estimates of \(\psi\) to its true value in 250 simulations per
scenario (the original analysis performed 500). The true value was
calculated by simulating the full data for \(500\,000\) individuals (the
original analysis used one million) with rules \(a_0\) and \(a_1\)
applied deterministically. Estimates of \(\psi\) were obtained by
estimating the survival curves \(S_a(t)\) using two estimators: the
inverse probability weighted Kaplan-Meier estimator (WKM) and the
Aalen-filtered WKM (AWKM). These survival estimators are detailed in the
following section.

@@

To estimate survival, we applied extensions of the widely-known
Kaplan-Meier (KM) estimator for survival (Kaplan and Meier 1958). First,
we review the estimator of Xie and Liu (2005), an extension of the KM
estimator where units are weighted by the inverse probability of
treatment. The standard KM estimator requires counting up the number of
cases \(c^0_{a}(t)\) that occurred in interval \((t - 1, t]\) and the
number of units at risk \(r^0_{a}(t)\) in that interval at all event
times \(t\). Assuming cancer status was assessed at the end of regular
intervals \(t = 1, \ldots, K\), we have: \[\begin{aligned}
c^0_{a} (t)
& = \sum_i^n \Ind{Y_i(t) = 1} \times \Ind{Y_i(t - 1) = 0} \times \Ind{\bar A_i (t) = \bar a(t)} \\
r^0_{a} (t)
& = \sum_i^n \Ind{Y_i(t - 1) = 0} \times \Ind{\bar A_i (t) = \bar a(t)}.
\end{aligned}\] The standard survival estimator is \[\hat S^0_{a}(t) =
\begin{cases}
1 & \text{if } t < t_1 \\
\prod_{j \le t} \left(1 - \frac{c^0_{a}(j)}{r^0_{a}(j)}\right) & \text{if } t \ge t_1 \\
\end{cases}\] where \(t_1\) is the first event time.

In observational studies, survival contrasts estimated using the
standard KM estimator are biased for the true causal survival contrast.
However, if conditional ignorability and positivity are attained, the
inverse probability weighted KM (WKM) estimator of Xie and Liu (2005)
yields unbiased estimates of the true causal survival curve. The WKM
estimator augments the standard KM estimator by weighting units at time
\(t\) by \(w_{i, a}(t)\) the inverse probability of treatment:
\[\begin{aligned}
c^w_{a} (t)
& = \sum_i^n w_{i, a}(t) \times \Ind{Y_i(t) = 1} \times \Ind{Y_i(t - 1) = 0} \times \Ind{\bar A_i (t) = \bar a(t)}\\
r^w_{a} (t)
& = \sum_i^n w_{i, a}(t) \times \Ind{Y_i(t - 1) = 0} \times \Ind{\bar A_i (t) = \bar a(t)}
\end{aligned}\] The WKM survival estimator for rule \(a\) is
\[\hat S^w_{a}(t) =
\begin{cases}
1 & \text{if } t < t_1 \\
\prod_{j \le t} \left(1 - \frac{c^w_{a}(j)}{R^w_{a}(j)}\right) & \text{if } t \ge t_1 \\
\end{cases}\] where \(t_1\) is the first event time.

Finally, to account for (uninformative) left filtering, we applied the
Aalen filter, which considers only the units at time \(t\) for which the
outcome is observed: \[\begin{aligned}
c_{a} (t)
& = \sum_i^n w_{i, a}(t) \times \Ind{Y_i(t) = 1} \times \Ind{Y_i(t - 1) = 0} \times \Ind{\bar A_i (t) = \bar a(t)} \times \Ind{t \ge R_i} \\
r_{a} (t)
& = \sum_i^n w_{i, a}(t) \times \Ind{Y_i(t - 1) = 0} \times \Ind{\bar A_i (t) = \bar a(t)} \times \Ind{t \ge R_i}
\end{aligned}\] The Aalen-filtered WKM (AWKM) estimator for rule \(a\)
is \[\hat S_{a}(t) =
\begin{cases}
1 & \text{if } t < t_1 \\
\prod_{j \le t} \left(1 - \frac{c_{a}(j)}{r_{a}(j)}\right) & \text{if } t \ge t_1 \\
\end{cases}\] where \(t_1\) is the first event time.

In the full data, the WKM and AWKM estimators are equivalent, and
identification is achieved under positivity and sequential ignorability
assumptions: \[\begin{aligned}
Y^*_{a, \bar d = 0} (t')   & \indep A(t) \mid W,\ \bar A(t - 1) = \bar a(t - 1),\ D(t - 1) = 0,\ N(t) = 1\\
Y^*_{a, \bar d = 0} (t') & \indep D(t) \mid W,\ D(t - 1) = 0,\ Y^*(t - 1) = 0,\ \bar A(t - 1) = \bar a(t - 1)
\end{aligned}\] for all times \(t'\ge t\), and
\[0 < \Prob{ A(t) = 1 \mid W,\ \bar A(t - 1) = \bar a(t - 1),\ D(t - 1) = 0,\ N(t) = 1} < 1\]
\[0 < \Prob{D(t) = 0 \mid W,\ D(t - 1) = 0,\ Y^*(t - 1) = 0,\ \bar A(t - 1) = \bar a(t - 1)} < 1.\]
